Autonomous Oceanographic Sampling Designs using Excursion Sets for Multivariate Gaussian random fields

### 2020-06-05 main_new.tex
updates

Note that this strategy gives a sequential conditional version of
expression (\ref{eq:eibv}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is conditional to this data. A similar closed-form calculation
for EIBV is hence applicable in Eq.
\eqref{critSEQ} using the
updated GP model from stage $s-1$. Once the data is collected for the
best design at stage $s$, the GP model is updated again. The mean $\bbm^{s}$ and
covariance matrix $\Sigma^{s}$ are used to compute the next design at
stage $s+1$, and so on.

Even though this myopic strategy is non-anticipative, it still gives a
reasonable approach for creating designs in many
applications. Moreover, it is easily implemented onboard an AUV,
using an efficient approach for data updating the GP model and the
calculation of the closed-form EIBV expressions for each
subsequent trajectory.

@@

We now extend the myopic strategy to a look-ahead strategy, which is
optimal when one can gather only two more stages of measurements. In
addition to the next stage of measurements $\by^{j,s}$, this
look-ahead strategy, anticipates the subsequent design indexed $j_2$ with
data $\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j} \left\{ E_{\by_{d^{j,s}}|\mathcal{Y}^{s-1}} \left( \mbox{argmin}_{j_2}  V_{j_2} \right) \right\}, \\
V_{j_2} & = & \int_{\mathcal{M}} E_{\by_{d^{j_2,s+1}}|\mathcal{Y}^{j,s}} \left( p(\x;\mathcal{Y}^{j,s+1}) ( 1-p(\x;\mathcal{Y}^{j,s+1}) ) \right) d\mes(u), \nonumber \\
    p(\x;\mathcal{Y}^{j,s+1}) &=& P(Z(\x) \in T |\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by_{d^{j,s}},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}\}$ represent
the sets of data variables, and the expectations are calculated with respect to
the conditional densities given data at that stage, meaning we have two levels of intermixed minimizations and integrals.
We solve the
first expectation by Monte Carlo sampling of data $\by_{d^{j,s}}$ from its
conditional distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV from
Section \ref{sec:ESEP}.

@@

We now study the properties of
the different static and sequential survey designs in a realistic
simulated case that emulates the spatial properties of a river plume interacting with coastal water.

@@

A GP model is used to represent the bivariate random fields of
temperature and salinity. At location $\x \in \mathcal{M}$, the
bivariate distribution is:
\begin{equation}\label{gp_bar}
\begin{bmatrix}Z_1(\x) \\
Z_2(\x) \end{bmatrix}
\sim N \left( 
\begin{bmatrix} \mu_{1}(\x)\\
\mu_{2}(\x)
\end{bmatrix},\begin{bmatrix}
\sigma_{1}^2 & \sigma_{1} \sigma_2 \gamma \\
\sigma_{1} \sigma_2 \gamma  & \sigma_{2}^2 
\end{bmatrix}
\right).
\end{equation}
We specify the mean by 
$\mu_{1}(\x)=5.8 + 0.065 u_{\mbox{West}}$ and
$\mu_{2}(\x)=29 + 0.1 u_{\mbox{West}}$, which means that the expected levels of both temperature and salinity increase linearly in the western direction.
This situation
mimics that of a river mouth opening from the south, with the water masses pulled to the east by the Coriolis force. 
The variances are set to $\sigma_1=0.25$ and $\sigma_2=0.6$, while the correlation is $0.6$. 
The spatial correlation is of separable Matern type (see e.g. \cite{Cressie:11}) so that
\begin{equation}\label{v}
\small
\mbox{Corr} 
\left(
\begin{bmatrix}
    Z_1(\x) \\
    Z_2(\x) 
    \end{bmatrix},
    \begin{bmatrix}
    Z_1(\x') \\
    Z_2(\x') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi h)\exp (-\phi h),
\end{equation}
where $h=\sqrt{||\x-\x'||^2}$ and $\phi=0.3$ indicates an effective correlation range of about $1200$ m.

In the accompanying Python code, these modeling assumptions can be generalized to anisotropic covariance and changing variance levels across the spatial domain. One can hence easily run the code to see if different models result in other sampling designs. Anisotropy and non-stationary variance are both relevant for the setting with river plumes, but in practice this requires more parameters to be specified. With extensive data and prior knowledge, one could also possibly fit and estimate parameters of more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}, but that is outside the scope of the current paper.

We discretize the spatial domain $\mathcal{M}$ to a set of $n$ grid
locations $\mathcal{M}_g = \{\x_i, i=1,\ldots,n \}$, where each cell
has area $\Delta$; the same grid is used for the waypoint graph setting the
possible design locations. The EIBV is approximated by sums over all grid cells.

@@

\begin{figure}[!ht]
  \centering
  \subfigure[Simulated temperature.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_temp.pdf}\label{fig:true_temp}}
  \hfill
  \subfigure[Simulated salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_sal.pdf}\label{fig:true_sal}}
  \hfill
  \subfigure[Excursion set given the realization of temperature and salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_true.pdf}\label{fig:ESet}}
  \hfill
  \subfigure[Excursion probabilities given collected data using the
  `static\_north" strategy.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_posterior.pdf}\label{fig:Eprob}}
  \caption{\ref{fig:true_temp} and \ref{fig:true_sal} show
    a realisation of the simulated temperature and salinity fields. The associated ES is in \ref{fig:ESet}, while 
    \ref{fig:Eprob} shows the estimated EPs after
    performing data collection along the static N-S survey line.} 
\label{fig:realisations}
\end{figure}

\subsection{Static and Sequential Sampling Designs}\label{sec:sampling_designs}

@@

For a fixed survey length, a closed-form expression for the EIBV is available as in Eq. \eqref{eq:eibv}. However, for the sequential approaches this is not the
case. For comparison the properties are therefore evaluated using
Monte Carlo sampling over several replicates of realizations from the model while conducting simulated sequential surveys for each
one. Fig. \ref{fig:Eprob} shows the conditional EP, given data
gathered along the north-south survey lines in these displays. In the
Monte Carlo replicates, such results are repeatedly computed to
approximate the EIBV. We also compare predictive performance
measured by root mean square error (RMSE) for temperature and salinity
estimates as well as also the variance reduction in these two variables. It is
important to note that the objective function used by the agent is
focused on reducing the EIBV, but we nevertheless expect that
we will achieve good predictive performance for criteria such as RMSE
as well. Another non-statistical criteria that is important for
practical purposes is the computational time needed for the strategy,
as this will impact the performance for an embedded system.

@@

Each strategy is conducted on an equilateral grid as shown in
Fig. \ref{fig:wp_graph}. The sequential sampling agent starts at the
center East-West coordinate at the southern end of the domain (node
53). The AUV moves along edges in the waypoint graph while the vehicle
gathers measurements. The data is assimilated into the GP model before
an evaluation of the next node to sample is conducted at the end of
the edge.

@@

\begin{figure}[h!]
  \centering
  % \subfigure[Excursion set variance $E_{\by}(p[1-p])$.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \subfigure[IBV.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \hfill
  \subfigure[RMSE between estimated field and truth.]{\label{fig:avg_rmse}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_RMSE.pdf}}
  \hfill 
  \subfigure[Explained variance $\bR^{2}$.]{\label{fig:avg_r2}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_R2.pdf}}
  \hfill 
  \subfigure[Time used to do inference.]{\label{fig:avg_time}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_Time.pdf}}
\caption{Simulation results from 100 replicate simulations for 10
  sampling choices/stages on the grid.} 
\label{fig:sim_results}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \subfigure[Look-ahead]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_look-ahead.pdf}\label{fig:avg_look-ahead}}
  \hfill
  \subfigure[Myopic]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_myopic.pdf}\label{fig:avg_myopic}}
  \hfill
  \subfigure[Naive]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_naive.pdf}\label{fig:route_naive}}
  \hfill
  \subfigure[Static]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_static.pdf}\label{fig:route_static}}
  \hfill
  \caption{The overlaid route choices (strategies), superimposed on
    the 100 replicate surveys with 10 sampling choices/stages.}
\label{fig:route_choices}
\end{figure}

Fig. \ref{fig:avg_time} shows the computational effort: the
\textit{naive} strategy is on par with the static designs, while the
\textit{myopic} strategy is slower. The \textit{look-ahead} is even
slower, reaching levels that are nearly impractical for execution on a
vehicle. Some pruning of the graph is performed to improve the
computational speed, such as ruling out repeated visitations and
back-and-forth routes. Some of the intermediate results are also
stored for longer planning horizons. Further pruning of branches or
inclusion of other heuristics could be included to make the strategy
faster. Then again, the inclusion of heuristics is likely a
contributing factor for the \textit{look-ahead} strategy failing to
outperform the \textit{myopic} strategy.

@@

We studied the sensitivity of the results by modifying the input parameters to have different correlations between temperature and salinity, standard deviations, and spatial correlation range.  In all runs,  the \textit{myopic} and \textit{look-ahead} strategies perform the best in terms of realized IBV, and much better than \textit{naive}. The \textit{look-ahead} strategy seems to be substantially better than the \textit{myopic} only for very small initial standard deviation or very large spatial correlation range. The \textit{static\_north} design continues to be the best static design for IBV, while \textit{static\_zigzag} is the best design for the other predictive performance measures, especially so with large spatial correlation range. We also ran simulation studies with only temperature data, and for realistic correlation levels between temperature and salinity the IBV results are not much worse when only temperature data are available. In addition to the comparison made in Table \ref{tab:sim_rhoab}, the current setting includes spatial correlation and this likely reduces the additional influence of having bivariate data. However, it seems that having temperature data alone does a substantially worse job in terms of explained variance

@@

\subsection{Model Specification}
\label{sec:exp_modeling}
The statistical model parameters were specified based on a short
preliminary survey where the AUV made an initial transect to determine the trends in environmental conditions and correlation structures. Based on the
initial data, the trend parameters were estimated by linear
regression, where both temperature and salinity are assumed to
increase linearly, going west from the river mouth. Next, the
residuals from the regression analysis were analyzed to study the fit
of the GP model and to specify the covariance parameters.

@@

\begin{table}[!h]
\centering
\begin{tabular}{lrr}
\toprule
Parameter & Value & Source\\
\midrule
\rowcolor{Gray}
Cross correlation temp. and sal. & 0.5 & AUV observations\\
Temp. variance &  0.20 & AUV observations (variogram)\\
\rowcolor{Gray}
Sal. variance &  5.76 & AUV observations (variogram)\\
Corr. range  & 0.15 km & AUV observations (variogram)\\
\rowcolor{Gray}
River temp. $T_{river}$ & $10.0\,^{\circ}\mathrm{C}$ & AUV observations\\
Ocean temp. $T_{ocean}$ & $11.0\,^{\circ}\mathrm{C}$ & AUV observations\\
\rowcolor{Gray}
River sal. $S_{river}$ & $14.0$ g/kg & AUV observations\\
Ocean sal. $S_{ocean}$ & $22.0$ g/kg & AUV observations\\
\rowcolor{Gray}
Threshold temp. & $10.5\,^{\circ}\mathrm{C}$ & $(T_{ocean}-T_{river})/2+T_{river}$\\
Threshold sal. & $18.0$ g/kg & $(S_{ocean}-S_{river})/2+S_{river}$\\
\rowcolor{Gray}
\bottomrule
\end{tabular}
\caption{Model and threshold parameters from an initial AUV
  survey. Observations were taken across the front while crossing from
  fresh, cold river water to saline and warmer ocean waters.}
\label{tab:experiment_param}
\end{table}

@@

\begin{figure*}[!h]
\centering
\subfigure[Survey map]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/alt_map.eps}\label{fig:map}}
\hspace{0.3cm}
\subfigure[Temperature tracks]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/auv.pdf}\label{fig:res_both}}

@@

%\subfigure[ES for Survey 1]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_1.pdf}\label{fig:res3}}\hspace{0.4cm}
%\subfigure[ES for Survey 2]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_4.pdf}\label{fig:res4}}
\caption{Results from mapping the Nidelva river. \ref{fig:map} Show an
  overview of the survey area overlaid with the AUV path in black and
  dashed line. Note the shaded region indicating the usual frontal
  region. \ref{fig:res_both} Presents the collected temperature data
  as colored trails. Note waypoint 5 (WP5) which indicate where the
  two surveys deviate. \ref{fig:res1} and \ref{fig:res2} Shows the
  collected salinity data (as a colored trail) overlaid on the final
  EP, which indicate the AUVs statistical impression of the front. For
  both missions the temperature and salinity data correspond with the
  EP front indication.}
\label{fig:results}
\end{figure*}

@@

Two missions, Surveys 1 and 2, were run successively from 11:00 AM to
01:00 PM, with a short break in between. The resulting path of the
selected waypoints are shown in the map in Fig. \ref{fig:map}, both
within the expected frontal region (shaded pink). The recorded
temperatures are shown as colored trails in Fig. \ref{fig:res_both},
clearly indicating the temperature difference between fjord and river
waters. The salinity data are then shown separately, overlaid with the
estimated EP for each survey in Fig. \ref{fig:res1} and
\ref{fig:res2}.

Both surveys successfully estimated and navigated the separation zone,
crossing the frontal boundary multiple times. As conditions changed
slightly between the two deployments, the resulting path (after
waypoint 5) is shown to deviate. Survey 1 continued northwards,
tracking the north-eastern part of the front, while Survey 2 turned
west, mapping the south-western part.

The final predictions of the front location, represented by conditional EPs in
Fig. \ref{fig:res1} and \ref{fig:res2} as dashed lines, correspond
with one another. In both surveys they yield a picture of the front being a bit to the west in the southern parts of the domain and gradually bending off toward the north east. The amount of exploration done by Survey 1
is greater than Survey 2. In Survey 1, the AUV obtained more detail by
going north from waypoint 5, while Survey 2, coming close to the survey area borders
in the south-western corner, obtained a poorer understanding of the
northern parts. A look-ahead strategy might identify and
discourage such choices.

@@

This work builds on a multidisciplinary effort combining statistical
modeling with robotic surveying techniques for oceanographic
applications. We show how observation practices can gain efficiency
and accuracy from the development of statistical techniques to achieve
more effective spatial monitoring. We further demonstrate the
opportunities available for real-time multivariable spatial
data-gathering and analysis onboard autonomous sensing platforms,
which statisticians can exploit to create general-purpose toolkits for
similar applications.

In particular, we derive and show results for characterizing phenomena
connected to the properties of water masses. The characterization of uncertainties in random sets is extended with new results for the EIBV uncertainty
reduction achieved by the spatial sampling design. This is provided in closed-form for the situation with a static
design, and then extended to the adaptive situation. The sequential
derivations provide new insights into efficient applications of
adaptive sampling, as demonstrated in our application of these models
with an autonomous vehicle in Trondheim fjord.

The study did not consider any temporal effects, which would be
relevant on a larger time scale. We consider the extension to
spatio-temporal modeling as future work, and envision that
advection-diffusion equations could be useful in this kind of modeling
\citep{sigrist2015stochastic}. For more complex oceanographic
phenomena, the methods will need to be extended to non-Gaussian
phenomena, possibly feature-based mixtures of GPs which could still be
run onboard and augmented by dynamical models.

>>>>

Note that this strategy gives a sequential conditional version of
expression (\ref{eq:eibv}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is conditional to this data. A similar closed-form
calculation for EIBV is hence applicable in Eq.  \eqref{critSEQ}
\kc{Strangely in my TeX compilation this points to the earlier
  definition of $d^{*,s}$. Seems like some index is messed up in the
  sources; all indexes should perhaps be checked?} using the updated
GP model from stage $s-1$. Once the data is collected for the best
\kc{??}  design at stage $s$, the GP model is updated again. The mean
$\bbm^{s}$ and covariance matrix $\Sigma^{s}$ are used to compute the
next design at stage $s+1$, and so on.

Even though this myopic strategy is non-anticipative, it still gives a
reasonable approach for creating designs in many
applications. Moreover, it is easily implemented, in our case, onboard
an AUV, using an efficient data update of the GP model and the
calculation of the closed-form EIBV expressions for each subsequent
trajectory.

@@

We now extend the myopic strategy to a look-ahead strategy, which is
optimal when one can gather only two more stages of
measurements. \kc{How can you be sure of this?} In addition to the
next stage of measurements $\by^{j,s}$, this look-ahead strategy,
anticipates the subsequent design indexed $j_2$ with data
$\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j} \left\{ E_{\by_{d^{j,s}}|\mathcal{Y}^{s-1}} \left( \mbox{argmin}_{j_2}  V_{j_2} \right) \right\}, \\
V_{j_2} & = & \int_{\mathcal{M}} E_{\by_{d^{j_2,s+1}}|\mathcal{Y}^{j,s}} \left( p(\x;\mathcal{Y}^{j,s+1}) ( 1-p(\x;\mathcal{Y}^{j,s+1}) ) \right) d\mes(u), \nonumber \\
    p(\x;\mathcal{Y}^{j,s+1}) &=& P(Z(\x) \in T |\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by_{d^{j,s}},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}\}$
represent the sets of data variables, and the expectations are
calculated with respect to the conditional densities given data at
that stage, implying that we have two levels of intermixed
minimizations and integrals.  We solve the first expectation by Monte
Carlo sampling of data $\by_{d^{j,s}}$ from its conditional
distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV from Section
\ref{sec:ESEP}.

@@

We now study the properties of the different static and sequential
survey designs in a realistic case that emulates the spatial
properties of a river plume interacting with coastal water.

@@

A GP model is used to represent the bivariate random fields of
temperature and salinity. At location $\x \in \mathcal{M}$, the
bivariate distribution is:
\begin{equation}\label{gp_bar}
\begin{bmatrix}Z_1(\x) \\
Z_2(\x) \end{bmatrix}
\sim N \left( 
\begin{bmatrix} \mu_{1}(\x)\\
\mu_{2}(\x)
\end{bmatrix},\begin{bmatrix}
\sigma_{1}^2 & \sigma_{1} \sigma_2 \gamma \\
\sigma_{1} \sigma_2 \gamma  & \sigma_{2}^2 
\end{bmatrix}
\right).
\end{equation}
We specify the mean by $\mu_{1}(\x)=5.8 + 0.065 u_{\mbox{West}}$ and
$\mu_{2}(\x)=29 + 0.1 u_{\mbox{West}}$, which means that the expected
levels of both temperature and salinity increase linearly in the
western direction.  This situation mimics that of a river mouth
opening from the south, with the water masses pulled to the east by
the Coriolis force \kc{This seems overly specific to the Nidelva use
  case. Can it be generalized, after all that is the overall intent of
  this manuscript?}.  The variances are set to $\sigma_1=0.25$ and
$\sigma_2=0.6$, while the correlation is $0.6$.  The spatial
correlation is of separable Matern type (see e.g. \cite{Cressie:11})
so that
\begin{equation}\label{v}
\small
\mbox{Corr} 
\left(
\begin{bmatrix}
    Z_1(\x) \\
    Z_2(\x) 
    \end{bmatrix},
    \begin{bmatrix}
    Z_1(\x') \\
    Z_2(\x') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi h)\exp (-\phi h),
\end{equation}
where $h=\sqrt{||\x-\x'||^2}$ and $\phi=0.3$ indicates an effective
correlation range of about $1200$ m.

In the accompanying Python code, these modeling assumptions can be
generalized to anisotropic covariance and changing variance levels
across the spatial domain. One can hence, easily run the code to see
if different models result in other sampling designs. Anisotropy and
non-stationary variance are both relevant for the setting with river
plumes, but in practice this requires more parameters to be
specified. With extensive data and prior knowledge, one could also
possibly fit and estimate parameters of more complex multivariate
spatial covariance functions
\citep{gneiting2010matern,genton2015cross}, but that is outside the
scope of the current paper.

We discretize the spatial domain $\mathcal{M}$ to a set of $n$ grid
locations $\mathcal{M}_g = \{\x_i, i=1,\ldots,n \}$, \kc{is the index
  correct? TeX shows it as u or g for $\mathcal{M}$} where each cell
has area $\Delta$; the same grid is used for the waypoint graph for
possible design locations. The EIBV is approximated by sums over all
grid cells.

@@

\begin{figure}[!h]
  \centering
  \subfigure[Simulated temperature.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_temp.pdf}\label{fig:true_temp}}
  \hfill
  \subfigure[Simulated salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/true_sal.pdf}\label{fig:true_sal}}
  \hfill
  \subfigure[Excursion set given the realization of temperature and salinity.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_true.pdf}\label{fig:ESet}}
  \hfill
  \subfigure[Excursion probabilities given collected data using the
  `static\_north" strategy.]{\includegraphics[height = 0.40\textwidth]{Figures/sim/es_ts_posterior.pdf}\label{fig:Eprob}}
  \caption{\ref{fig:true_temp} and \ref{fig:true_sal} show a
    realization of the simulated temperature and salinity fields. The
    associated ES is in \ref{fig:ESet}, while \ref{fig:Eprob} shows
    the estimated EPs after performing data collection along the
    static N-S survey line.}
\label{fig:realisations}
\end{figure}

\subsection{Static and Sequential Sampling Designs}
\label{sec:sampling_designs}

@@

For a fixed survey length, a closed-form expression for the EIBV is
available as in Eq. \eqref{eq:eibv}. However, for the sequential
approaches this is not the case. For comparison the properties are
therefore evaluated using Monte Carlo sampling over several replicates
of realizations from the model while conducting simulated sequential
surveys for each one. Fig. \ref{fig:Eprob} shows the conditional EP,
given data gathered along the north-south survey lines in these
displays. In the Monte Carlo replicates, such results are repeatedly
computed to approximate the EIBV. We also compare predictive
performance measured by root mean square error (RMSE) for temperature
and salinity estimates as well as also the variance reduction in these
two variables. It is important to note that the objective function
used by the agent \kc{First of 'agent'. Define it or change it
  to 'robot'? Note further uses below.} is focused on reducing the
EIBV, but we nevertheless expect that we will achieve good predictive
performance for criteria such as RMSE as well. Another non-statistical
criteria that is important for practical purposes is the computational
time needed for the strategy, as this will impact the performance for
an embedded system.

@@

Each strategy is conducted on an equilateral grid as shown in
Fig. \ref{fig:wp_graph}. The sequential sampling agent starts at the
center East-West coordinate at the southern end of the domain (node 53
\kc{Maybe color this node differently and modify the caption?}). The
AUV moves along edges in the waypoint graph while the vehicle makes
measurements. The data is assimilated into the GP model before an
evaluation of the next node to sample is conducted at the end of the
edge.

@@

\begin{figure}[h!]
  \centering
  % \subfigure[Excursion set variance $E_{\by}(p[1-p])$.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \subfigure[IBV.]{\label{fig:avg_ev}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_EV.pdf}}
  \hfill
  \subfigure[RMSE between estimated field and truth.]{\label{fig:avg_rmse}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_RMSE.pdf}}
  \hfill 
  \subfigure[Explained variance $\bR^{2}$.]{\label{fig:avg_r2}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_R2.pdf}}
  \hfill 
  \subfigure[Computational time for inferencing.\kc{I see only 
    lines associated with 4 variables showing in the graph. Where is
    static\_north and static\_east?}]{\label{fig:avg_time}\includegraphics[height=0.49\textwidth]{Figures/sim/avg_Time.pdf}} 
\caption{Simulation results from 100 replicate simulations for 10
  sampling choices/stages on the grid. \kc{Vertical lines show large
    variation in replicate results.}}  
\label{fig:sim_results}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \subfigure[Look-ahead]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_look-ahead.pdf}\label{fig:avg_look-ahead}}
  \hfill
  \subfigure[Myopic]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_myopic.pdf}\label{fig:avg_myopic}}
  \hfill
  \subfigure[Naive]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_naive.pdf}\label{fig:route_naive}}
  \hfill
  \subfigure[Static]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_static.pdf}\label{fig:route_static}}
  \hfill
  \caption{The overlaid route choices (strategies), superimposed on
    the 100 replicate surveys with 10 sampling choices/stages. \kc{I
      would reorder these figurs above based on the order the material
      was discussed -- Static, naive, myopic and then Look-ahead. Also
      do you really need a key associated with each subfigure, given
      the caption? An abridged analysis as in the text, might also be
      in order here in the caption. Also refer to
      Fig. \ref{fig:realisations} as the reference, where the
      temp/salinity fields are shown.}}
\label{fig:route_choices}
\end{figure}

Fig. \ref{fig:avg_time} shows the computational effort: the
\textit{naive} strategy is on par with the static designs, while the
\textit{myopic} strategy is slower. The \textit{look-ahead} is even
slower, reaching levels that are nearly impractical for execution on a
vehicle. Some pruning of the graph is performed to improve the
performance, such as ruling out repeated visitations and
back-and-forth routes. Some of the intermediate results are also
stored for longer planning horizons. Further pruning of branches or
inclusion of other heuristics could be included for better
performance. Then again, the inclusion of heuristics is likely a
contributing factor for the \textit{look-ahead} strategy failing to
outperform the \textit{myopic} strategy.

@@

We studied the sensitivity of the results by modifying the input
parameters to have different correlations between temperature and
salinity, standard deviations, and spatial correlation range.  In all
runs, the \textit{myopic} and \textit{look-ahead} strategies perform
the best in terms of realized IBV, and much better than
\textit{naive}. The \textit{look-ahead} strategy seems to be
substantially better than the \textit{myopic} design only for very
small initial standard deviation or very large spatial correlation
range. \textit{static\_north} continues to be the best static design
for IBV, while \textit{static\_zigzag} is the best design for the
other predictive performance measures, especially so with large
spatial correlation range. We also ran simulation studies with only
temperature data, and for realistic correlation levels between
temperature and salinity, the IBV results are not much worse when only
temperature data are available. In addition to the comparison made in
Table \ref{tab:sim_rhoab}, the current setting includes spatial
correlation and this likely reduces the additional influence of having
bivariate data. However, it seems that having temperature data alone
does a substantially worse job in terms of explained variance

@@

\subsection{Model Specification}
\label{sec:exp_modeling}

The statistical model parameters were specified based on a short
preliminary survey where the AUV made an initial transect to determine
the trends in environmental conditions and correlation
structures. Based on the initial data, the trend parameters were
estimated by linear regression, where both temperature and salinity
are assumed to increase linearly, going west from the river
mouth. Next, the residuals from the regression analysis were analyzed
to study the fit of the GP model and to specify the covariance
parameters.

@@

\begin{table}[!h]
\centering
\begin{tabular}{lrr}
\toprule
Parameter & Value & Source\\
\midrule
\rowcolor{Gray}
Cross correlation temp. and sal. & 0.5 & AUV observations\\
Temp. variance &  0.20 & AUV observations (variogram)\\
\rowcolor{Gray}
Sal. variance &  5.76 & AUV observations (variogram)\\
Corr. range  & 0.15 km & AUV observations (variogram)\\
\rowcolor{Gray}
River temp. $T_{river}$ & $10.0\,^{\circ}\mathrm{C}$ & AUV observations\\
Ocean temp. $T_{ocean}$ & $11.0\,^{\circ}\mathrm{C}$ & AUV observations\\
\rowcolor{Gray}
River sal. $S_{river}$ & $14.0$ g/kg & AUV observations\\
Ocean sal. $S_{ocean}$ & $22.0$ g/kg & AUV observations\\
\rowcolor{Gray}
Threshold temp. & $10.5\,^{\circ}\mathrm{C}$ & $(T_{ocean}-T_{river})/2+T_{river}$\\
Threshold sal. & $18.0$ g/kg & $(S_{ocean}-S_{river})/2+S_{river}$\\
\rowcolor{Gray}
\bottomrule
\end{tabular}
\caption{Model and threshold parameters from an initial AUV
  survey. Observations were taken across the front while crossing from
  fresh, cold river water to saline and warmer ocean waters. \kc{Might
  be good to expand the terms and not have abbreviations in the table
  above; there's plenty of space.}}
\label{tab:experiment_param}
\end{table}

@@

\begin{figure*}[!h]
\centering
\subfigure[AUV survey area]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/alt_map.eps}\label{fig:map}}
\hspace{0.3cm}
\subfigure[Temperature tracks]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/auv.pdf}\label{fig:res_both}}

@@

%\subfigure[ES for Survey 1]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_1.pdf}\label{fig:res3}}\hspace{0.4cm}
%\subfigure[ES for Survey 2]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_4.pdf}\label{fig:res4}}
\caption{Results from mapping the Nidelva river, Trondheim, Norway
  over two survey missions. \ref{fig:map} shows an overview of the
  survey area overlaid with the AUV path in black and dashed
  line. Note the shaded region indicating a typical frontal
  region. \ref{fig:res_both} shows the collected temperature data as
  colored trails. Note waypoint 5 (WP5) which indicates where the two
  surveys diverge. \ref{fig:res1} and \ref{fig:res2} shows the
  collected salinity data overlaid on the final EP, which indicate the
  AUVs statistical impression of the front. For both missions the
  temperature and salinity data correspond with an indication of the
  EP front. \kc{Might want to indicate when the surveys (dates) were
    conducted so people can understand the variability in the EP with
    respect to time.}}
\label{fig:results}
\end{figure*}

@@

Two surveys missions (1 and 2), were run successively from 11:00 AM to
01:00 PM, with a short break in between. The resulting path of the
selected waypoints are shown in the map in Fig. \ref{fig:map}, both
within the expected frontal region (shaded pink). The recorded
temperatures are shown as colored trails in Fig. \ref{fig:res_both},
clearly indicating the temperature difference between fjord and
riverine waters. The salinity data are then shown separately, overlaid
with the estimated EP for each survey in Fig. \ref{fig:res1} and
\ref{fig:res2}.

Both surveys successfully estimated and navigated the separation zone,
crossing the frontal boundary multiple times. As conditions changed
slightly between the two surveys, the resulting path (after waypoint
5) is shown to deviate. Survey 1 continued northwards, tracking the
north-eastern portion of the front, while Survey 2 turned west,
mapping the south-western region.

The final predictions of the front location, represented by
conditional EPs in Fig. \ref{fig:res1} and \ref{fig:res2} as dashed
lines, correspond with one another. In both surveys they yield a
picture of the front being to the west in the southern portions of the
region and gradually bending off toward the north east. The amount of
exploration done by Survey 1 is greater than Survey 2. In Survey 1,
the AUV obtained more detail by going north from waypoint 5, while
Survey 2, coming close to the survey area borders in the south-western
corner, obtained a poorer understanding of the northern parts. A
look-ahead strategy might identify and discourage such choices.

@@

This work builds on a multidisciplinary effort combining statistical
modeling with robotic surveying techniques for oceanographic
applications. We show how observation practices can gain efficiency
and accuracy from the development of statistical techniques to achieve
more effective spatial monitoring. We further demonstrate the
opportunities available for real-time multivariable spatial
data-gathering and analysis onboard autonomous platforms, which
statisticians can exploit to create general-purpose toolkits for
similar applications.

In particular, we derive and show results for characterizing phenomena
connected to the properties of water masses. The characterization of
uncertainties in random sets is extended with new results for the EIBV
uncertainty reduction achieved by spatial sampling design. This is
provided in closed-form for the situation with a static design, and
then extended to the adaptive situation. The sequential derivations
provide new insights into efficient applications of adaptive sampling,
as demonstrated in our application.

The study did not consider any temporal effects, which would be
relevant on a larger time scale. We consider the extension to
spatio-temporal modeling as future work, and envision that
advection-diffusion equations could be useful in this kind of modeling
\citep{sigrist2015stochastic}. For more complex oceanographic
phenomena, the methods will need to be extended to non-Gaussian
phenomena, possibly feature-based mixtures of GPs which could still be
run onboard and augmented by dynamical models. \kc{How about talking
  about removing any grids or waypoints altogether?}

The spatial-statistical design criterion, building on random sets is
relevant in our setting with different water properties, and we show
mathematical generality beyond the EIBV which is convenient to
implement for onboard robotic planning. Such criteria could also be
useful in other oceanographic settings as for algal-blooms, anoxic
zones or open water fronts \cite{costa19}. Of course, other criteria
are also relevant for instance, hybrid or multi-attribute criteria
that could balance goals of exploration and exploitation in this
situation. Equally, such techniques have significant use cases in
downstream decision-making, with policy makers and regulators who need
to make difficult decisions related to fish farming, other marine
resource operations. Value of information analysis (VOI)
\citep{Eidsvik:15} could be used in a similar vein as the IBV in the
current work, as both analytic methods explore when, where and which
information is likely to result in improved decision-making.
